---
title: "FDZ Checks"
author: "Annegret Rucker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FDZ Checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates how data providers can use the `check_all()` function from the `eatFDZ` package to assess and improve the quality of their data sets prior to submission to the [Research Data Centre](https://fdz.iqb.hu-berlin.de/de/) (FDZ) at the [Institute for Educational Quality Improvement](https://www.iqb.hu-berlin.de/) (IQB).
For data to be reusable in a research data center context, it must meet high standards regarding structure, consistency, documentation, and data protection. In particular, data sets should be FAIR (findable, accessible, interoperable, and reusable), well-documented, and prepared in a way that allows for efficient and reliable secondary use. To support this process, `check_all()` provides a comprehensive set of automated checks that help identify potential issues early on. This vignette explains how to run `check_all()`, how to interpret its output, and how to use the results to systematically improve data set quality.
Overall, the following aspects are covered by `check_all()`:

- Valid file naming conventions
- Consistent and valid variable names
- Proper encoding of metadata
- Unique and non-missing identifier variables
- Presence of variable labels
- Correct specification of missing value labels and ranges
- Identification of undocumented or irregular missing values
- Detection of low-frequency categories (statistical disclosure control)
- Consistency between data set and external documentation (e.g., codebooks)

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```
The setup for this illustration is the following:
``` {r path}
## if necessary, install eatGADS and eatFDZ
#install.packages("eatGADS")
#remotes::install_github("beckerbenj/eatFDZ")

# load eatGADS
library(eatGADS)
library(eatFDZ)

# path to example data set
sav_path <- system.file("extdata", "example_data.sav", package = "eatFDZ")
```
`eatGADS` is the R package used by the FDZ at IQB for data cleaning purposes. It is available on [CRAN](https://cran.r-project.org/package=eatGADS) and extensively documented via vignettes (e.g., the [Comprehensive Data Cleaning Guide](https://beckerbenj.github.io/eatGADS/articles/data_cleaning.html) or the [FDZ Data Cleaning Guide](https://beckerbenj.github.io/eatFDZ/articles/FDZ_Data_Cleaning.html)).

For illustrative purposes, a small, artificial SPSS example data set of secondary school students in Berlin is used. The data set is intentionally flawed and messy. The example data set is part of the `eatFDZ` package and installed alongside. 

## Quickstart
```{r load}
dat <- import_spss(sav_path)
str(dat$dat) # for a small overview in the variables of the data set
```
The data set is loaded via the `import_spss()` command. The data set contains 14 variables with 10 observations, including demographic variables such as age and sex of the students, their family situation and selected grades.

Note that the data set is imported as a `GADSdat` object. For detailed explanations on the corresponding object structure and technical details, we refer readers to the `eatGADS` [documentation](https://beckerbenj.github.io/eatGADS/articles/import_spss.html). 


