---
title: "FDZ Data Cleaning"
author: "Annegret Rucker, Benjamin Becker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FDZ Data Cleaning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates how the [Research Data Centre](https://www.iqb.hu-berlin.de/fdz) (FDZ) at the [Institute for Educational Quality Improvement](https://www.iqb.hu-berlin.de/) (IQB) cleans and tidies data for secondary usage purposes. 
For secondary usage, data is expected to be of high quality, traceable, comprehensible, and anonymized. Furthermore, meaningful metadata should enable secondary users to understand the data. 
For this purpose, operations such as recoding of identifier variables, emptying variables, changing labels, creating grouping variables, and many more are performed and therefore discussed in this vignette. Overall, the following aspects should be adressed when cleaning data:

- Appropriate missing tags
- Non-informative identifier variables
- Transforming strings to labeled numeric variables
- Anonymization of variables
- Adding variable and value labels

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```
The setup for this illustration is the following:
``` {r path}
## if necessary, install eatGADS and eatFDZ
#install.packages(eatGADS)
#remotes::install_github("beckerbenj/eatFDZ")

# load eatGADS
library(eatGADS)

# path to example data set
sav_path <- system.file("extdata", "example_data.sav", package = "eatFDZ")
```
`eatGADS` is the R package used by the FDZ at IQB for data cleaning purposes. It is available on [CRAN](https://cran.r-project.org/package=eatGADS) and extensively documented via vignettes (e.g., the [Comprehensive Data Cleaning Guide](https://beckerbenj.github.io/eatGADS/articles/data_cleaning.html)).

For illustrative purposes, a small, artificial SPSS example data set of secondary school students in Berlin is used. The data set is intentionally flawed and messy. The example data set is part of the `eatFDZ` package and installed alongside. 

## Loading the data
```{r load}
dat <- import_spss(sav_path)
str(dat$dat) # for a small overview in the variables of the data set
```
The data set is loaded via the `import_spss()` command. The data set contains 14 variables with 10 observations, including demographic variables such as age and sex of the students, their family situation and selected grades.

Note that the data set is imported as a `GADSdat` object. For detailed explanations on the corresponding object structure and technical details, we refer readers to the `eatGADS` [documentation](https://beckerbenj.github.io/eatGADS/articles/import_spss.html). 


## Define missings semi-automatically
A crucial meta data property for SPSS data are **missing tags**.
Sometimes not all missing values are correctly tagged as missing. Via `eatGADS`, missing tags can be corrected semi-automatically. This can either be done based on wordings of existing value labels (`checkMissings()`) or based on a value range of existing value labels (`checkMissingsByValues()`).

`checkMissings()` tests whether a set of specified regular expressions occurs in the value labels and compares the corresponding missings tags. For instance the variable `books` contains a value label `omitted`, which has not been tagged as missing.
```{r checkMissings}
extractMeta(dat, vars = c("books")) # usually, we wouldn't know yet that this variable contains these value labels. This is for demonstration purposes to see the before and after effect of the function
missinglabels <- paste("missing", "unknown",
                       "omitted", "unclear answer",
                       sep = "|")
dat1 <- checkMissings(dat, missingLabel = missinglabels)
extractMeta(dat1, vars = c("books"))
```
`checkMissings()` has now automatically tagged the value `-99` with the label `omitted` as missing (`miss`).

`checkMissingsByValues()` checks whether value labels exist for a specified range of numeric values and whether missing tags are provided accordingly:
```{r checkMissingsByValues}
# tbd: Use case is missing!
dat2 <- checkMissingsByValues(dat1, missingValues = -50:-99)
```


## Define missing values by hand
At times, only individual missings need to be tagged. `changeMissings()` tags missing values as missings by specifying them under `value` and marking them as `miss` under `missings.`
```{r changeMissings}
extractMeta(dat2, vars = c("info"))
dat3 <- changeMissings(dat2, varName = "info", value = c("-97", "-98", "-99"), missings = c("miss", "miss", "miss"))
extractMeta(dat3, vars = c("info"))
```
The values `-97`, `-98` and `-99` are now tagged as `miss`.


## Recode IDs
To ensure that the IDs remain the same across multiple subsets and describe the same person, they are recoded using a template. First, a path for the template is created.
```{r IDrecode, eval = FALSE}
f <- tempfile()
ID_recode_template <- file.path("f") # use your own file path instead of the tempfile
```
Then there are two options: Creating a new template (option 1 - for the first subset) or applying an existing template (option 2 - for the following subsets).

### Option 1: IDs recoded with creation of a template as .csv
`autoRecode()` creates a recoding template for the variable `ID`, saved as a .csv file, and in the course of this a new variable with the variable suffix `_FDZ` and an explanatory label suffix.
```{r autoRecode, eval = FALSE}
dat3 <- autoRecode(dat3, var = "ID", var_suffix = "_FDZ", csv_path = ID_recode_template, label_suffix = " (recoded by FDZ)")
```
### Option 2: IDs recoded with application of a template (e.g. created by a previously prepared, different data set).
Already existing templates can be used for following subsets. `read.csv()` imports the template, which can be applied with `autoRecode()`
```{r autoRecode2, eval = FALSE}
template <- read.csv(ID_recode_template)
dat3 <- autoRecode(dat3, var = "ID", suffix = "_FDZ", csv_path = ID_recode_template, template = template,
                   label_suffix = "(rekodiert FDZ)") # Apply template and overwrite
```

`removeVars()` deletes variables such as the old ID variable from the data set.
```{r removeVars, eval = FALSE}
dat3 <- removeVars(dat3, vars = c("ID"))
```


## Convert string variables to numeric variables
For (semi)open variables there are two procedures. If it is possible (e.g., for a small data set) to go through all responses to check whether there are no strong privacy violations or insults included, the variables are converted into numeric ones and then emptied in the Scientific Use Files. Thereby, the information remain accessible on a meta-level, but can no longer be assigned to a person. If there are too many responses to check or privacy cannot be granted, the variable is emptied entirely in the Scientific Use Files.
To convert the string variables into numeric ones, all string variables are identified first.
```{r docall}
all_types <- do.call(rbind, lapply(namesGADS(dat3), function(nam) data.frame(varName = nam, type = class(dat3$dat[[nam]]))))
all_types[all_types$type == "character", ]
```
In this case the variables `ID_name`, `info`, `subjfav` and `birth` were recognized as string variables. In the next step they are converted.

### Option 1: Convert all string variables automatically (overwrite variables)
`multiChar2fac()` converts the string variables identified in advance into numeric variables using a for loop.
```{r allTypes}
charNames <- all_types[all_types$type == "character", "varName"]
for(charName in charNames) {
  dat4 <- multiChar2fac(dat3, vars = charName, var_suffix = "", label_suffix = "")
}
```
### Option 2: Convert string variable individually
`multiChar2fac()` converts the string variables named in `vars` individually into numeric variables. If the option `var_suffix` is used, a new variable is created.
```{r multiChar2fac, eval=FALSE}
dat4 <- multiChar2fac(dat4, vars = c("ID_name", "info", "subjfav", "birth"), var_suffix = "_FDZ", label_suffix = "")
```
If the `var_suffix` option is left blank, the existing variable is overwritten.
```{r multiChar2fac2, eval=FALSE}
dat4 <- multiChar2fac(dat4, vars = c("ID_name", "info", "subjfav", "birth"), var_suffix = "", label_suffix = "")
```


## Empty variables already in IZA
Variables that may not even be offered via a remote computing system due to strong data protection concerns are emptied for all data sets. `emptyTheseVariables()` empties the listed variables and adds an explanation to the variable label on why the information was removed.
```{r emptyTheseVariables}
extractMeta(dat4, vars = c("info", "ID_name"))
dat4_empty <- c("info",
                "ID_name")
dat5 <- emptyTheseVariables(dat4,
                            vars = dat4_empty,
                            label_suffix = " (Zur Anonymisierung geleert (FDZ))")
extractMeta(dat5, vars = c("info", "ID_name"))
```
`dat4` corresponds to the data set, `vars` lists the variables to be emptied and `label_suffix` automatically adds the explanation to the variable label indicating that the data has been emptied for anonymization reasons.


## Change variable and value labels
Variable and value labels help us with the traceability and comprehensibility of a data set. In `ChangeVarLabels()` the variable to be changed is selected under `varName` and with `varLabel` the corresponding label is assigned. `ChangeValLabels()` works the same way, except that the corresponding value is selected, too.

1. Changing variable labels
```{r changeVarLabels}
extractMeta(dat5, vars = c("school"))
dat6 <- changeVarLabels(dat5, varName = "school", varLabel = "School type")
extractMeta(dat6, vars = c("school"))
```
2. Changing value labels
``` {r changeValLabels}
extractMeta(dat6, vars = c("books"))
dat7 <- changeValLabels(dat6, varName = "books", value = 5, valLabel = "201-500 books")
extractMeta(dat7, vars = c("books"))
```


## Anonymization: Form grouping variable
For anonymization reasons, we partially form grouping variables for items with a case number <6. To do this, we create a new _FDZ variable by "cloning" the variable we want to group with `cloneVariable()`. For `varName` we enter the original variable name, for `new_varName` we enter the variable name including the appendix "_FDZ". The variable label is automatically adjusted with the option `label_suffix`.
With `recodeGADS()` the old values (`oldValues`) are replaced by new ones (`newValues`). With the help of `changeValLabels()` the value labels of the new groupings are changed.
```{r recodeGADS}
extractMeta(dat7, vars = c("age"))
dat8 <- cloneVariable(dat7, varName = "age", new_varName = "age_FDZ", label_suffix = " (Zur Anonymisierung aggregiert (FDZ))")

dat8 <- recodeGADS(dat8, varName = "age_FDZ", oldValues = c(13, 14, 15, 17),
                  newValues = c(13, 13, 15, 15))
dat8 <- changeValLabels(dat8, varName = "age_FDZ", value = c(13, 15), valLabel = c("13-14", "15-17"))
extractMeta(dat8, vars = c("age_FDZ"))
```
You can also perform grouping for multiple variables using a loop. For this we give the variables a temporary name consisting of the variable name and the suffix "_FDZ". `varname` represents each individual variable, which was summarized before in `gr_var`. The individual steps (`cloneVariable()`, `recodeGADS()`,`changeValLabels()`) are therefore performed for each of the variables in turn.
```{r recodeGADS_loop}
extractMeta(dat8, vars = c("grade_math"))
dat9 <- dat8
gr_var <- c("grade_math", "grade_germ", "grade_eng")
for (varname in gr_var) {
  tempnewname <- paste0(varname, "_FDZ")
  dat9 <- cloneVariable(dat9, varName = varname,
                        new_varName = tempnewname, label_suffix = " (Zur Anonymisierung aggregiert (FDZ))")
  dat9 <- recodeGADS(dat9, varName = tempnewname,
                      oldValues = c(1:6),
                      newValues = c(rep(1,4),rep(2,2)))
  dat9 <- changeValLabels(dat9, varName = tempnewname, value = c(1, 2),
                           valLabel = c("passed", "not passed"))
}
extractMeta(dat9, vars = c("grade_math_FDZ"))
```
Now we have created two groups. The first group includes the students with grades from 1-4, which means they have passed. In the other group, there are the students with grades 5 and 6, which means they have not passed. These groups were created for all three subjects Math, German and English at once.


## Creating the version variable
To indicate the state of the data set, each data set is provided with a version variable. This is created using `createVariable()` and placed at the beginning of the record by `relovateVariable()` so that it has a consistent, visible place.
```{r VersionsVar}
version_name <- "Version_v1_2023_07"
dat10 <- createVariable(dat9, varName = version_name)
dat10 <- relocateVariable(dat10, var = version_name, after = NULL)
```


## Change variable order
To rearrange the variables of a data set, the function `orderLike()` helps, in which the variables are listed in their new order.
```{r orderLike}
dat11 <- orderLike(dat10, newOrder = c("Version_v1_2023_07", "ID", "ID_name", "info", "home", "birth", "age", "age_FDZ", "sex", "siblings", "books", "school", "subjfav", "grade_math", "grade_math_FDZ", "grade_germ", "grade_germ_FDZ", "grade_eng", "grade_eng_FDZ"))
```
To sort in individual variables, such as newly created grouping variables behind their emptied original variable, `relocateVariable()` can be used.
```{r relocateVariable}
dat11 <- relocateVariable(dat11, var = "age_FDZ", after = "age")
```
