---
title: "`check_all()`: Running automated data quality and documentation checks"
author: "Annegret Rucker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`check_all()`: Running automated data quality and documentation checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates how data providers can use the `check_all()` function from the `eatFDZ` package to assess and improve the quality of their data sets prior to submission to the [Research Data Centre](https://fdz.iqb.hu-berlin.de/en/) (FDZ) at the Institute for Educational Quality Improvement (IQB).

For data to be reusable in a research data center context, it must meet high standards regarding structure, consistency, documentation, and data protection. In particular, data sets should be FAIR (findable, accessible, interoperable, and reusable), well-documented, and prepared in a way that allows for efficient and reliable secondary use. To support this process, `check_all()` provides a comprehensive set of automated checks that help identify potential issues early on. This vignette explains how to run `check_all()`, how to interpret its output, and how to use the results to systematically improve data set quality.

## What `check_all()` does (and what it does not do)

**`check_all()` does:**
- run a set of automated checks on a `.sav` file (imported as a `GADSdat` object)
- return a structured set of result tables (plus an overview)
- help you spot typical issues, such as invalid variable names, missing labels, missing IDs, missing tags, and potential low-frequency disclosure risks

**`check_all()` does not:**
- automatically fix your data
- replace project-specific, expert review of the data and documentation
- guarantee anonymization; the SDC check should be interpreted as a *risk indicator*


## Checks covered by `check_all()`

Overall, the following aspects are covered by `check_all()`:

- Valid file naming conventions
- Consistent and valid variable names
- Proper encoding of metadata
- Unique and non-missing identifier variables
- Presence of variable labels
- Correct specification of missing value labels and ranges
- Identification of undocumented or irregular missing values
- Detection of low-frequency categories (statistical disclosure control)
- Optional: consistency between data set and external documentation (e.g., codebooks)

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Setup
To run the examples in this vignette, install and load the `eatFDZ` package. For installation from GitHub, you can use `remotes::install_github("beckerbenj/eatFDZ")`.

Then load the package and set the path to your data set.
``` {r path}
library(eatFDZ)

# path to example data set
sav_path <- system.file("extdata", "example_data.sav", package = "eatFDZ")
```
For illustrative purposes, a small, artificial SPSS example data set of secondary school students in Berlin is used. The data set is intentionally flawed and messy. The example data set is part of the `eatFDZ` package and installed alongside. 


## Quickstart: create an Excel check report

```{r quickstart}
out <- check_all(sav_path = sav_path)
report_path <- file.path(tempdir(), "check_report.xlsx")
write_check_report(out, file_path = report_path)
```
For real projects, instead of `tempdir()` save reports to a dedicated folder.

Open the generated Excel file at `report_path`.


## How to read the Excel report

The Excel report is designed for fast navigation:

- `out$Overview`: a compact summary indicating whether each check passed, found issues, or was not run
- additional list elements: one result table per check where issues were found (each a `data.frame`)

- **Passing**: no issues found, there won't be a corresponding result table

- **Issues detected**: the result table contains at least one row

- **Not tested**: the check was skipped (e.g., documentation checks when `pdf_path` is not provided)

### Typical workflow

1. Open the Excel report.
2. Start with **Overview** to identify which checks need attention.
3. For each check with **Issues detected**, review the corresponding sheet.
4. Fix the issues in the source data and/or metadata.
5. Re-run `check_all()` and regenerate the report until the Overview is mostly *Passing*.


## Common customisations

`check_all()` offers a small set of parameters that you can adapt to your project context.

### Specify the identifier variable (`idVar`)

By default, `check_all()` uses the first variable in the data set as the identifier. If your ID variable is not the first one (or you have multiple IDs), set `idVar` explicitly:

```{r idvar, eval=FALSE}
out <- check_all(
  sav_path = sav_path,
  idVar = "student_id"
)
```

### Define project-specific missing value rules (`missingRange`, `missingRegex`)

If your project uses a defined range of missing codes (e.g., `-97`, `-98`, `-99`) you can specify this via `missingRange`:

```{r missingrange, eval=FALSE}
out <- check_all(
  sav_path = sav_path,
  missingRange = -99:-97
)
```

If missing values are indicated via labels (possibly multilingual), adapt the regular expression `missingRegex`:

```{r missingregex, eval=FALSE}
out <- check_all(
  sav_path = sav_path,
  missingRegex = "missing|omitted|not reached|no answer|keine angabe"
)
```

### Restrict statistical disclosure control checks (`sdcVars`)

By default, `check_all()` runs SDC checks on **all variables**. To focus on sensitive variables or key domains, you can specify `sdcVars`:

```{r sdcvars, eval=FALSE}
out <- check_all(
  sav_path = sav_path,
  sdcVars = c("school_id", "class_id", "migration_background")
)
```


## Optional: documentation checks via `pdf_path`

If you provide `pdf_path`, `check_all()` will run documentation checks (e.g., whether variables are referenced in the codebook/documentation).

```{r docu, eval=FALSE}
pdf_path <- "path/to/codebook.pdf"

out <- check_all(
  sav_path = sav_path,
  pdf_path = pdf_path
)
```

If `pdf_path` is not provided, the documentation check is reported as **Not tested** in the Overview.


## Batch processing for multiple data sets

If you need to check multiple `.sav` files (e.g., for several cohorts or waves), you can run `check_all()` in a loop and write one Excel report per file.

```{r batch, eval=FALSE}
data_dir <- "path/to/sav_files"
report_dir <- "reports"
dir.create(report_dir, showWarnings = FALSE, recursive = TRUE)

sav_files <- list.files(data_dir, pattern = "\\.sav$", full.names = TRUE, recursive = FALSE)

for (f in sav_files) {
  message("Checking: ", f)

  # Create a stable output name
  base <- sub("\\.sav$", "", basename(f), ignore.case = TRUE)
  report_path <- file.path(report_dir, paste0(base, "_check_report.xlsx"))

  # Ensure one failing file does not stop the entire batch
  out <- tryCatch(
    check_all(sav_path = f),
    error = function(e) {
      message("  ERROR: ", conditionMessage(e))
      return(NULL)
    }
  )

  if (!is.null(out)) {
    write_check_report(out, file_path = report_path)
    message("  Report written: ", report_path)
  }
}
```

### Searching subfolders

If your `.sav` files are stored in subfolders, set `recursive = TRUE`:

```{r batch-recursive, eval=FALSE}
sav_files <- list.files(data_dir, pattern = "\\.sav$", full.names = TRUE, recursive = TRUE)
```


## Troubleshooting

### `docu_check` shows “Not tested”
- **Cause:** you did not provide `pdf_path`
- **Fix:** call `check_all(..., pdf_path = "path/to/codebook.pdf")`

### Many encoding-related findings
- **Cause:** non-UTF-8 characters or inconsistent encoding in names/labels/metadata
- **Fix:** check how the `.sav` file was created/exported; if needed, set the `encoding` argument (and ensure your toolchain uses UTF-8 consistently)

### Unexpected missing or duplicate IDs
- **Cause:** the wrong ID variable was used (default is the first variable)
- **Fix:** set `idVar` explicitly

### Many SDC flags
- **Cause:** small sample sizes, fine-grained categories, or non-sensitive variables being checked
- **Fix:** interpret SDC flags as *risk indicators*; consider restricting `sdcVars` to variables relevant for disclosure risk

### Reports are saved “somewhere I cannot find”
- **Cause:** you wrote to `tempdir()`, which is session-specific
- **Fix:** save reports into a dedicated project folder (e.g., `reports/`) and print the path after writing

### Batch processing is slow
- **Cause:** SDC checks and documentation checks can be time-consuming
- **Fix:** restrict `sdcVars`; run docu checks only when needed; process files individually if required


## Further resources
If you want to continue with systematic cleaning and preparation after running `check_all()`, see:

[FDZ Data Cleaning Guide](https://beckerbenj.github.io/eatFDZ/articles/FDZ_Data_Cleaning.html)

[Comprehensive Data Cleaning Guide](https://beckerbenj.github.io/eatGADS/articles/data_cleaning.html)
