---
title: "`check_all()`: Running automated data quality and documentation checks"
author: "Annegret Rucker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`check_all()`: Running automated data quality and documentation checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates how data providers can use the `check_all()` function from the `eatFDZ` package to assess and improve the quality of their data sets prior to submission to the [Research Data Centre](https://fdz.iqb.hu-berlin.de/en/) (FDZ) at the Institute for Educational Quality Improvement (IQB).

For data to be reusable in a research data center context, it must meet high standards regarding structure, consistency, documentation, and data protection. In particular, data sets should be FAIR (findable, accessible, interoperable, and reusable), well-documented, and prepared in a way that allows for efficient and reliable secondary use. To support this process, `check_all()` provides a comprehensive set of automated checks that help identify potential issues early on. This vignette explains how to run `check_all()`, how to interpret its output, and how to use the results to systematically improve data set quality.

Overall, the following aspects are covered by `check_all()`:

- Valid file naming conventions
- Consistent and valid variable names
- Proper encoding of metadata
- Unique and non-missing identifier variables
- Presence of variable labels
- Correct specification of missing value labels and ranges
- Identification of undocumented or irregular missing values
- Detection of low-frequency categories (statistical disclosure control)
- Consistency between data set and external documentation (e.g., codebooks)

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Setup
To run the examples in this vignette, install and load the `eatFDZ` package.  

For installation from GitHub, you can use `remotes::install_github("beckerbenj/eatFDZ")`.
``` {r path}
library(eatFDZ)

# path to example data set
sav_path <- system.file("extdata", "example_data.sav", package = "eatFDZ")
```
For illustrative purposes, a small, artificial SPSS example data set of secondary school students in Berlin is used. The data set is intentionally flawed and messy. The example data set is part of the `eatFDZ` package and installed alongside. 


## Quickstart
```{r quickstart}
out <- check_all(sav_path = sav_path)

report_path <- file.path(tempdir(), "check_report.xlsx")
write_check_report(out, file_path = report_path)

# report_path
```
Open the Excel file at `report_path`.

Hier möchte ich erklären, was "Passing", "Issues detected" und "Not tested" bedeuten.

- **Passing**: corresponding result table is empty (no issues found)

- **Issues detected**: result table contains at least one row

- **Not tested**: check was skipped (e.g., documentation checks without `pdf_path`)


## Exporting and interpreting the output
```{r export}
report_path <- file.path(tempdir(), "check_report.xlsx")
write_check_report(out, report_path)
```
Hier möchte ich zeigen, wie man den Report exportiert  und was darin enthalten ist (Überblick und einzelne Prüfungen - Reiter entsteht nur, wenn "Issued detected" bei der jeweiligen Prüfung).

Erwähnen, dass statt tempdir() der eigene Projektordner genutzt werden soll.


## Common customisations
Hier möchte ich mit erklären, wie man idVar setzt (wenn ID nicht 1. Variable ist), die Missing Definitionen an die Datensätze angepasst werden, der SDC Scope eingeschränkt wird

## Optional: documentation checks via `pdf_path`
Was wird hier geprüft (variables refernced in codebook)? 

In der Vignette erstmal als "eval=FALSE" bis ich Beispiel-Codebook erstellt habe.

„If pdf_path is not provided, docu_check is reported as Not tested.“


## Troubleshooting
- docu_check Not tested → pdf_path fehlt
- many encoding findings → check encoding / metadata
- unexpected ID findings → specify idVar
- SDC flags many categories → interpret as risk indicator; restrict sdcVars if needed
- where are outputs saved → show paths / tempdir() vs user-defined folder
- Why does batch processing take long? → SDC / docu checks can be time-consuming; run per file / restrict scope.
- recursive = FALSE: Datensätze sind in Unterordnern

## Further resources
Wenn Datensatz im Anschluss an die Prüfung aufbereitet werden soll, dann sind hier gute Referenzen:

[FDZ Data Cleaning Guide](https://beckerbenj.github.io/eatFDZ/articles/FDZ_Data_Cleaning.html)

[Comprehensive Data Cleaning Guide](https://beckerbenj.github.io/eatGADS/articles/data_cleaning.html)
