---
title: "`check_all()`: Preparing data for quality review and submission"
author: "Annegret Rucker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`check_all()`: Preparing data for quality review and submission}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Purpose of this vignette

This vignette explains how data providers can use the `check_all()` function from the `eatFDZ` package to review and improve their data sets before submitting them to the [Research Data Centre](https://fdz.iqb.hu-berlin.de/en/) (FDZ) at the Institute for Educational Quality Improvement (IQB).

`check_all()` performs automated checks of SPSS `.sav` files and their metadata. These checks help identify common problems related to:

- file and variable naming,
- identifier variables,
- variable and value labels,
- missing-value definitions,
- data structure,
- potential statistical disclosure risks,
- character variables,
- consistency between a data set and its documentation.

> **Scope:** The function supports the technical and documentation-related preparation of data for FAIR-aligned reuse by identifying potential issues that require further review.


---

## What `check_all()` does — and what it does not do

`check_all()`:

- imports an SPSS `.sav` file internally as a `GADSdat` object,
- runs a collection of automated checks,
- returns an overview and detailed result tables,
- helps identify issues before the data are submitted,
- supports a systematic and traceable review process.

`check_all()` does not:

- modify or clean the data,
- decide automatically whether every finding is an error,
- replace project-specific knowledge,
- guarantee anonymisation or data protection compliance.


---

## Before you start

**Before running the checks, you should know:**

- the path to the `.sav` file,
- which variables serve as identifiers,
- which numeric codes represent missing values,
- which terms are used in missing-value labels,
- where the corresponding codebook or documentation is stored.

> **Version consistency:** The data set and documentation should refer to the same version.


---

## Setup

Install `eatFDZ` from GitHub if it is not already installed. Then load the package.

```{r install, eval = FALSE}
remotes::install_github("beckerbenj/eatFDZ")
library(eatFDZ)
```

For the examples in this vignette, we use a small artificial SPSS data set included in the package:

```{r example-path}
sav_path <- system.file("extdata", "example_data.sav", package = "eatFDZ")
```

The example data set contains intentionally introduced problems so that the different checks produce illustrative findings.

> **For your own data:** Replace `sav_path` with the path to your `.sav` file.


---

## Quickstart: create an Excel report

**Use this section if you want to run the checks immediately.**

The following code runs the checks and writes the results to an Excel file:

```{r quickstart, eval = FALSE}
out <- check_all(sav_path = sav_path)
report_path <- file.path("reports", "check_report.xlsx")
write_check_report(out, file_path = report_path)
report_path
```

**Next step:** Open the Excel file saved at `report_path` and start with the `Overview` sheet.

> **Important:** The file name is checked before the data are imported. An invalid file name may therefore produce a message or error before the remaining checks are performed.

A suitable file name should:

- identify the data set clearly,
- avoid umlauts, spaces, and problematic special characters,
- include a comprehensible version number where appropriate.


---

## Working through the report

The Excel report consists of an `Overview` sheet and additional sheets containing detailed findings.

The `Overview` indicates whether a check:

- found no issues,
- produced findings that require review,
- was not run.

> **Interpretation:** Not every reported finding is necessarily an error. Some findings may be expected because of the study design or data structure. Nevertheless, every finding should be reviewed.

### Recommended workflow

1. Open the `Overview` sheet.
2. Identify checks that produced findings.
3. Open the corresponding result sheets.
4. Determine whether each finding represents:
   - an error that must be corrected,
   - an expected feature of the data,
   - or an issue that requires further assessment.
5. Correct the source data or metadata where necessary.
6. Document findings that are intentional or cannot be changed.
7. Run `check_all()` again.
8. Generate a new report for the corrected data set.

The final data set, codebook, and check report should all describe the same version of the data.


---

## Adapting the checks to your project

**Use this section to adjust the checks to your data structure and project conventions.**

The default settings may not be appropriate for every data set. The following arguments can be adapted to the structure and conventions of your project.

### Identifier variables: `idVar`

By default, the first variable in the data set is used as the identifier variable.

If the relevant identifier is stored in another variable, specify it explicitly:

```{r idvar, eval = FALSE}
out <- check_all(
  sav_path = sav_path,
  idVar = "student_id"
)
```

Several identifier variables can be supplied as a character vector:

```{r multiple-ids, eval = FALSE}
out <- check_all(
  sav_path = sav_path,
  idVar = c(
    "school_id",
    "class_id",
    "student_id"
  )
)
```

Interpret duplicate identifiers in relation to the intended data structure. For example, a school ID is expected to occur multiple times in a student-level data set.

If observations are uniquely identified only by a combination of variables, verify the combined key separately.

### Missing-value range: `missingRange`

Projects often use a defined numeric range for missing codes, such as `-97`, `-98`, and `-99`.

Specify this range using `missingRange`:

```{r missingrange, eval = FALSE}
out <- check_all(
  sav_path = sav_path,
  missingRange = -99:-97
)
```

The check reports values within this range that occur in the data but are not defined as missing in the metadata.

> **Caution:** Do not change a value solely because it falls within the specified range. First verify whether it represents a missing value or a substantively valid response.

### Missing-value labels: `missingRegex`

`missingRegex` searches value labels for terms that indicate missing data.

Multiple terms can be separated using `|`:

```{r missingregex, eval = FALSE}
out <- check_all(
  sav_path = sav_path,
  missingRegex = paste(
    "missing",
    "omitted",
    "not reached",
    "no answer",
    sep = "|"
  )
)
```

Adapt the expression to the terminology and languages used in your project.

> **Tip:** Avoid overly broad search terms, as they may produce false positives.

### Variables for disclosure control: `sdcVars`

By default, the statistical disclosure control check is run on all variables.

To restrict the check to selected variables, use `sdcVars`:

```{r sdcvars, eval = FALSE}
out <- check_all(
  sav_path = sav_path,
  sdcVars = c(
    "school_id",
    "migration_background",
    "special_educational_need"
  )
)
```

Select variables based on their sensitivity, level of detail, and potential contribution to identification risks.

### Documentation: `pdf_path`

If a searchable PDF codebook is available, it can be included in the checks:

```{r docu, eval = FALSE}
out <- check_all(
  sav_path = sav_path,
  pdf_path = "path/to/codebook.pdf"
)
```

> **Note:** If `pdf_path` is not supplied, the documentation check is shown as not tested.

> **Requirement:** The PDF must contain searchable text. Variable names that appear only in images or scanned pages may not be detected.

### Complete example

A project-specific call may look as follows:

```{r complete-example, eval = FALSE}
out <- check_all(
  sav_path = "data/study_data_v1_0.sav",
  idVar = "student_id",
  missingRange = -99:-97,
  missingRegex = paste(
    "missing",
    "omitted",
    "not reached",
    "no answer",
    sep = "|"
  ),
  sdcVars = c(
    "school_id",
    "migration_background"
  ),
  pdf_path = "documentation/codebook_v1_0.pdf"
)

write_check_report(
  out,
  file_path = "reports/study_data_v1_0_check_report.xlsx"
)
```


---

## Reference: individual checks

**Use this section when a result sheet contains findings and you want to understand what they mean.**

The following sections explain how the individual result tables should be interpreted.

### Naming and encoding

<details>
<summary><code>lengthy_variable_names</code></summary>

Reports variable names containing more than 30 characters.

Long variable names may cause problems across statistical software and make analysis scripts difficult to read.

Review whether the names can be shortened systematically. The full meaning should be preserved in the variable label and documentation.

</details>

<details>
<summary><code>special_signs_variable_names</code></summary>

Reports variable names containing problematic characters or encodings.

Variable names should preferably contain only:

- letters,
- numbers,
- underscores.

For example:

- `größe` could be changed to `groesse`,
- `test.score` could be changed to `test_score`.

Changes must also be reflected in the documentation and processing scripts.

Functions such as `eatGADS::fixEncoding()` may help correct encoding problems.

</details>

<details>
<summary><code>special_signs_meta_data</code></summary>

Reports potentially problematic characters in variable labels or value labels.

Check whether the affected text is displayed correctly in:

- R,
- SPSS,
- exported files,
- the codebook.

Always apply encoding corrections to a working copy and verify the result afterwards.

</details>

### Identifier variables

<details>
<summary><code>missing_IDs</code></summary>

Reports observations with missing values in the specified identifier variables.

Missing IDs can prevent observations from being assigned, linked, or traced reliably.

Check whether:

- the ID was lost during data processing,
- an empty string or missing code was used,
- the observation should be included,
- the ID can be reconstructed from a reliable source.

Do not invent identifiers without a documented basis.

</details>

<details>
<summary><code>duplicate_IDs</code></summary>

Reports identifier values occurring more than once.

Possible causes include:

- duplicated rows,
- incorrect merges,
- an unsuitable identifier variable,
- a hierarchical data structure.

> **Interpretation:** A duplicate value is not necessarily incorrect. For example, school and class IDs normally occur repeatedly in student-level data.

Determine which variable or combination of variables is expected to identify an observation uniquely.

</details>

### Labels and missing values

<details>
<summary><code>missing_variable_labels</code></summary>

Reports variables without an informative variable label.

Variable labels should briefly describe what a variable measures or contains.

Unsuitable labels include:

- empty labels,
- placeholders such as `XXX`,
- labels that only repeat the variable name,
- undocumented internal processing notes.

Variable labels can be edited using functions such as `eatGADS::changeVarLabels()`.

</details>

<details>
<summary><code>missing_value_labels</code></summary>

Reports numeric values occurring in the data for which no value label is available.

> **Note:** Not every numeric variable requires value labels. Continuous variables such as age, scores, or time measurements may remain unlabelled.

For categorical variables, check whether:

- all valid categories are labelled,
- all missing codes are labelled,
- the labels are understandable without additional project knowledge.

Value labels can be edited using functions such as `eatGADS::changeValLabels()`.

</details>

<details>
<summary><code>empty_value_labels</code></summary>

Reports value labels that are empty or contain no meaningful description.

Check whether the value represents:

- a valid category requiring a label,
- a missing code requiring a description,
- an obsolete definition that should be removed.

</details>

<details>
<summary><code>missing_range_tags</code></summary>

Reports values within `missingRange` that are not defined as missing in the metadata.

For example, the value `-99` may have the label `omitted` but still be treated as a valid value.

First verify whether the value actually represents missing data. Missing-value definitions can then be corrected using functions such as `eatGADS::changeMissings()`.

</details>

<details>
<summary><code>missing_regex_tags</code></summary>

Reports values whose labels contain a term from `missingRegex` but whose values are not defined as missing.

Review the complete label and its substantive context. Search terms may also match valid categories unintentionally.

</details>

### Data structure

<details>
<summary><code>empty_variables</code></summary>

Reports variables containing no valid information, for example because all values are missing.

Possible explanations include:

- filter routing,
- data collection problems,
- failed imports,
- processing errors,
- variables intended for another subsample.

Remove variables without a documented purpose. Intentionally empty variables should be explained in the documentation.

</details>

<details>
<summary><code>constant_variables</code></summary>

Reports variables containing only one value across all observations.

Constant variables may be intentional, for example when all observations belong to the same country, wave, or experimental condition.

Check whether:

- variation was expected,
- values were overwritten or lost,
- the variable is required for merging or documentation.

Retain constant variables only when their purpose is clear.

</details>

### Data protection and character variables

<details>
<summary><code>statistical_disclosure_control</code></summary>

Reports categories with low frequencies that may indicate a disclosure risk.

Small frequencies are not automatically unacceptable. The actual risk depends on:

- the sensitivity of the information,
- possible combinations with other variables,
- whether persons or institutions may be identifiable,
- the intended access conditions.

Possible measures include:

- combining categories,
- removing sensitive information from a data version,
- using more restrictive access conditions,
- documenting a justified exception.

> **Important:** This automated check does not replace a complete statistical disclosure control assessment.

</details>

<details>
<summary><code>character_variables</code></summary>

Reports variables stored as character variables.

Character variables are not necessarily incorrect. However, text fields may:

- contain direct or indirect identifiers,
- include inconsistent spellings,
- be difficult to analyse,
- lack structured labels.

For every character variable, decide whether it:

- must remain a text variable,
- can be converted into a categorical variable,
- should be recoded,
- must be removed or emptied for data protection reasons.

Structured character variables can be converted using functions such as `eatGADS::multiChar2fac()`.

</details>

### Documentation

<details>
<summary><code>docu_check</code></summary>

Compares variable names in the data set with the supplied PDF documentation.

The `count` column indicates how often a variable name was found:

- `count == 0`: the variable name was not found,
- `count > 0`: the variable name was found at least once.

A missing match may mean that:

- the variable is not documented,
- the variable name differs between data and documentation,
- the variable appears only in an image,
- the PDF does not contain searchable text,
- the codebook and data set describe different versions.

> **Limitation:** The check only determines whether a variable name was found. It does not assess whether the substantive description is complete, correct, or understandable.

</details>
---

## When is a data set ready for submission?

From the perspective of the automated checks, a data set is well prepared when:

- all unexpected findings have been corrected,
- remaining findings have been reviewed and documented,
- identifiers are complete and appropriate for the intended data structure,
- categorical variables have informative variable and value labels,
- missing codes are labelled and defined correctly,
- empty and constant variables have been assessed,
- low-frequency sensitive categories have been reviewed,
- character variables have been checked deliberately,
- every variable is included in the relevant documentation,
- the data set, documentation, and report refer to the same version.

> **Good to know:** A completely passing `Overview` is a useful goal, but it may not be possible or substantively appropriate for every data set.

**The key requirement is that every reported finding has been reviewed and addressed in a transparent and traceable way.**


---

## Advanced: checking multiple data sets

Several `.sav` files can be processed in a loop. The following example writes one Excel report per data set:

```{r batch, eval = FALSE}
data_dir <- "path/to/sav_files"
report_dir <- "reports"

dir.create(report_dir, showWarnings = FALSE)

sav_files <- list.files(
  data_dir,
  pattern = "\\.sav$",
  full.names = TRUE
)

for (sav_file in sav_files) {
  report_file <- file.path(
    report_dir,
    sub("\\.sav$", "_check_report.xlsx", basename(sav_file))
  )

  tryCatch(
    write_check_report(
      check_all(sav_file),
      report_file
    ),
    error = \(e) {
      message(basename(sav_file), ": ", conditionMessage(e))
    }
  )
}
```

If files are stored in subfolders, add `recursive = TRUE` to `list.files()`.

Project-specific arguments such as `idVar`, `missingRange`, or `pdf_path` may differ between files. In that case, define the required settings separately for each data set.


---

## Troubleshooting

Start with the message shown in R and the corresponding sheet in the Excel report.

<details>
<summary><strong>The documentation check was not run</strong></summary>

**Possible cause:** No `pdf_path` was supplied.

**Solution:** Include the path to a searchable PDF:

```{r troubleshooting-docu, eval=FALSE}
check_all(
  sav_path = sav_path,
  pdf_path = "path/to/codebook.pdf"
)
```

</details>

<details>
<summary><strong>Unexpected missing or duplicate identifiers are reported</strong></summary>

**Possible cause:** The wrong identifier variable was used or the finding is expected because of the hierarchical data structure.

**Solution:** Specify `idVar` explicitly and review which variables or variable combinations should be unique.

</details>

<details>
<summary><strong>Many disclosure-control findings are reported</strong></summary>

**Possible cause:** The data set contains small samples, highly detailed categories, or variables that are not relevant for disclosure control.

**Solution:** Treat the findings as risk indicators and consider restricting `sdcVars` to relevant variables.

</details>

<details>
<summary><strong>Variable names are not found in the documentation</strong></summary>

**Possible causes:**

- the PDF contains scanned pages,
- variable names are written differently,
- the wrong documentation version was supplied.

**Solution:** Verify that the PDF contains searchable text and that the data and documentation refer to the same version.

</details>

<details>
<summary><strong>The report cannot be overwritten</strong></summary>

**Possible cause:** The Excel file is currently open or overwriting has not been enabled.

**Solution:** Close the file and, where appropriate, use the available overwrite option of `write_check_report()`.

</details>

---

## Further resources

> **Note:** `check_all()` identifies potential issues but does not modify the data set.

Guidance on cleaning and preparing data with `eatGADS` and `eatFDZ` is available in:

- [Data Curation at the FDZ at the IQB](https://beckerbenj.github.io/eatFDZ/articles/FDZ_Data_Cleaning.html)
- [Data Cleaning Guide with the R package eatGADS](https://beckerbenj.github.io/eatGADS/articles/data_cleaning.html)

---
